---
title: "Code-along, Week 02: *Billboard* Hot 100"
author: "Alex Homer"
date: "30 September 2021"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(scales)
library(lubridate)
library(emo)
#To install the "emo" package run:
# devtools::install_github("hadley/emo")
#You may need to install the "devtools" package first, in the normal way.
```

## Read data
The data are drawn from the "Tidy Tuesday" project: [2021 Week 38: Top 100 Billboard](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-09-14/readme.md) (credit: [Data.World](https://data.world/kcmillersean/billboard-hot-100-1958-2017#), via Sean Miller, [Billboard.com](https://www.billboard.com/) and Spotify).  We have very lightly modified the data by converting the `week_id` column to a week that R can recognise.

```{r read-data, message = FALSE}
chart_data <- readRDS("data/billboard.rds")
chart_data %>%
  summarise(
    first_week = min(chart_date),
    last_week = max(chart_date)
  )
```

The data span from 2nd August 1958 (which according to [Wikipedia](https://en.wikipedia.org/wiki/Billboard_Hot_100#History) was the first *Billboard* Hot 100 chart, although Wikipedia has the date a couple of days later) to 29th May 2021.

The full dataset has a row (i.e. an observation) for every week and chart position, so any song that was on the charts for more than one week appears multiple times.  For example, if we take [one particular song](https://www.youtube.com/watch?v=dQw4w9WgXcQ)...

```{r give-you-up, message = FALSE}
never_gonna <- chart_data %>%
  filter(song == "Never Gonna Give You Up") %>%
  arrange(chart_date) %>%
  select(song, performer, chart_date, week_position)

never_gonna %>% print(n = dim(never_gonna)[1]) #Print all rows
```

... we see that it was on the chart for many weeks.  This allows us to plot the chart fortunes of this song:

```{r let-you-down, message = FALSE}
never_gonna %>% ggplot(aes(x = chart_date, y = week_position)) +
  geom_line() +
  scale_x_date(
    labels = date_format("%d %b %Y"),
    expand = expansion(0.05, c(0,10))
  ) +
  scale_y_reverse(breaks = c(1, 10*(1:10))) + #Lower number = higher chart position
  labs(
    x = "Date",
    y = "Chart position",
    title = "I just wanna tell you how I'm charting",
    subtitle = "Gotta make you understand",
    caption = "Source: Data.World via Sean Miller, Billboard.com and Spotify"
  ) +
  coord_cartesian(ylim = c(100, 1)) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

Let's try a different song.

```{r one-republic}
chart_data %>% filter(song == "Counting Stars") %>%
  ggplot(aes(x = chart_date, y = week_position)) +
  geom_line() +
  scale_y_reverse()
```

`r emo::ji("bulb")` *This code chunk shows what we wrote during the live code-along.  Compare the output of this code to the output of the code above; see if you can deduce what the extra lines in the first version do.* `r emo::ji("bulb")`

Anyway, this is probably a bit "too much data" for our purposes in this code-along, so I've summarised the data so that each song appears only once, with some useful statistics for each song.  I've then put the results in a new .rds file, which we now import.

```{r read-data-2, message = FALSE}
song_data <- readRDS("data/songs.rds")
```

## Danceability
The data contain a score for "danceability" of the songs.  We can use this to find the most danceable songs...

```{r most-danceable-song}
song_data %>% arrange(desc(danceability)) %>%
  select(song, performer, danceability)
```

... as well as the least.

```{r least-danceable-song}
song_data %>% arrange(danceability) %>%
  select(song, performer, danceability)
```

I suspect that there's been some sort of data problem with the least danceable song, given that it has a score of exactly 0 (and if you [listen to it](https://www.youtube.com/watch?v=BQncPrWaxgI) it certainly doesn't seem like an extremely un-danceable song.  Let's investigate by looking at its values for some of the other scores.

```{r letter-to-an-angel}
song_data %>% filter(song == "A Letter To An Angel") %>%
  select(energy, acousticness, speechiness, liveness)
```

Maybe that is the real value, then?  Certainly "real data" seem to appear for the other values.  But if we do exclude it, then the least danceable song is "[An American Trilogy](https://www.youtube.com/watch?v=dQw4w9WgXcQ)" by Elvis Presley.

We should take this with a pinch of salt, though, because not all songs had a danceability value.

```{r dance-na}
song_data %>%
  summarise(dance_val_prop = sum(is.na(danceability))/n())
```

17.5% didn't, in fact.

What if we look at the average danceability by year?  First we need a "year" variable (we'll take the year a song first entered the charts as its "year", since we don't have data on release dates).

```{r average-danceability-year}
song_data <- song_data %>%
  mutate(song_year = year(first_appearance))

song_data %>%
  group_by(song_year) %>%
  summarise(average_dance = mean(danceability, na.rm = TRUE)) %>%
  ggplot(aes(x = song_year, y = average_dance)) +
  geom_line() +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Average danceability of songs",
    title = "Danceability seems to be weakly increasing"
  )
```

There seems to be a dip in danceability cntered on around 2010.  If we look at the least danceable songs, it's easy to blame this on one particular TV show...

```{r 2010-least}
song_data %>% filter(song_year == 2010) %>%
  arrange(danceability) %>%
  select(song, performer, danceability)
```

... but the *Glee* cast appear heavily in the most danceable songs too, including the most danceable, so that's probably unfair.

```{r 2010-most}
song_data %>% filter(song_year == 2010) %>%
  arrange(desc(danceability)) %>%
  select(song, performer, danceability)
```

Instead of doing the average, let's do a scatter plot of all songs' danceability against their first appearance in the chart.

```{r danceability-scatter, message = FALSE, warning = FALSE}
song_data %>%
  ggplot(aes(x = first_appearance, y = danceability)) +
  geom_point(alpha = 0.15) + #So we can see all the points
  labs(
    x = "First appearance in charts",
    y = "Danceability"
  ) +
  geom_smooth() + #Trend line
  theme_minimal()
```

Suddenly the 2010 dip is a lot less pronounced!  Although it still shows up in the trend line.

## Weeks on the chart

Let's go a little bit further than the year split, and assign each song to the decade in which it first appeared on the chart.

```{r decades}
song_data <- song_data %>%
  mutate(
    decade = cut(
      first_appearance,
      breaks = seq.Date(
        from = as.Date("1950/01/01"),
        to = as.Date("2030/01/01"),
        by = "10 years"
      ),
      labels = c("1950s", "1960s", "1970s", "1980s", "1990s",
                 "2000s", "2010s", "2020s"),
      right = FALSE
    )
  )
```

We can then count the number of songs that charted in each decade.

```{r n-per-decade}
count(song_data, decade)
#We could write this as
# song_data %>% count(decade)
#but it's a bit pointless using the pipe when there's only one step
```

Unsurprisingly there are fewer songs from the 1950s (because the chart started in 1958) and from the 2020s (because they haven't finished yet).  One thing that's noticeable, though, is that there were a lot more songs in the 1960s.  We can visualise that using a bar chart.

```{r bar-decade}
song_data %>% ggplot(aes(x = decade)) +
  geom_bar()
```

For each decade, we can plot a histogram of the number of weeks songs spent on the chart. ^[This is the total number of weeks, so for *any* song this could still increase if the song re-entered the chart.  The original data accounted for this by assigning "instances" for every time a song re-entered the charts: perhaps we should have grouped by each song *and instance*, rather than just each song?]  We do this using faceting.

```{r histogram-weeks}
song_data %>% ggplot(aes(x = weeks_on_chart)) +
  facet_wrap(~decade) +
  geom_histogram(breaks = 3*(0:29 + 0.5)) #Each block is three years
```

`r emo::ji("bulb")` *How could we make either of these plots prettier?* `r emo::ji("bulb")`

Two things are evident here:

* Firstly, songs are tending to spend longer on the charts as the decades go on---although from the 2000s onwards the number of songs spending extremely short amounts of time on the chart increases.
* Secondly, there's a strange peak at around 20 weeks, where we suddenly get lots of songs having that length of time on the charts.  This only appears from the 1990s.

You can see both features in the following graph, where, for each possible number of weeks spent on the chart from 1 to 30, we show what proportion of such songs spent exactly that number of weeks on the chart.

```{r bar-stack}
song_data %>% filter(weeks_on_chart <= 30) %>%
  ggplot(aes(x = weeks_on_chart, fill = decade)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Number of weeks on chart",
    y = "Proportion",
    fill = "Decade",
    caption = "Source: Data.World via Sean Miller, Billboard.com and Spotify",
    title = "In general, songs that spent longer on the charts tend to be newer...",
    subtitle = "... with something weird going on for songs that spent 1 and 20 weeks there!"
  ) +
  theme_minimal()
```

`r emo::ji("bulb")` *The data aren't really all that valid for songs that are still on the charts, because they haven't really dropped off the charts yet.  This will affect 100 songs in the 2020s section.  What could we do about that?  Does it matter?* `r emo::ji("bulb")`

The trends are pretty appearent here: if the song spent longer on the charts, it's more likely to be newer.  But what's going on at 1 and 20 weeks?

For the very short numbers of weeks, notice that a majority of songs that have spent exactly one week on the charts are from the 2010s or later, with also far more in the 2000s than in the 1990s or 1980s.  What changed about singles in that period?  Well, *Billboard* [started tracking downloads](https://en.wikipedia.org/wiki/Billboard_Hot_100#Digital_downloads,_online_streaming_and_bundles) for the Hot 100 in 2005.  Downloads (and now streaming, included since 2007) can create more volatile swings for songs, because there doesn't need to be a physical shipment of 45s/cassettes/CDs to make a song enter the charts.

The coloured bar graph makes it clear that the weird behaviour at 20 weeks affects particularly the data for songs spending *exactly* 20 weeks on the chart, and the histograms suggest that this started in the 1990s.  Such a sharp peak is unlikely to be the result of natural purchasing/streaming habits.  In fact, it can be explained by a [different policy change](https://en.wikipedia.org/wiki/Billboard_Hot_100#Recurrents) on the part of *Billboard*: "if a song has spent 20 weeks on the Hot 100 and fallen below position number 50", it is given "recurrent status", and removed from the chart. ^[Source of quote: [Wikipedia](https://en.wikipedia.org/wiki/Billboard_Hot_100#Recurrents).]  It's quite neat, I think, that we can see this clearly reflected in the data!